---
title: "Cast Vote Record Analysis - Cecil County, MD 2024 General"
author: "Earl F Glynn<br><small>watchdoglab.substack.com/</small>"
date: "<small>`r Sys.Date()`</small>"
output:
  html_document:
    code_download: true
    theme: cerulean
    toc: yes
    toc_depth:  3
    toc_float:
      collapsed:  yes
      smooth_scroll: yes
    number_sections: yes
    code_folding:  show
---

```{r setup, echo = FALSE}
# http://biostat.mc.vanderbilt.edu/wiki/Main/KnitrHtmlTemplate
require(Hmisc)    # provides knitrSet and other functions
knitrSet(lang = 'markdown',   # If using blogdown: knitrSet(lang='blogdown')
         fig.align = 'left',
         w = 6.5,
         h = 4.5,
         cache = FALSE)
```

`r hidingTOC(buttonLabel = "Outline")`

```{r startYourEngines, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  comment = NA)

time.1 <- Sys.time()
```

# Setup {.tabset .tabset-fade .tabset-pills}

## {.active}

## Constants

```{r}
EXCEL_LIMIT <- 2^20
```

## Packages

```{r Packages}
library(tidyverse)
library(lubridate)
```

Display formatting

```{r Display}
library(kableExtra)  # kable_styling
```

I/O

```{r IO}
library(readxl)      # read_xlsx
library(readr)       # write_csv
library(writexl)     # write_xlsx
```

 Graphics

```{r Graphics}
library(scales)      # comma, comma_format
```

Utilities

```{r Utilities}
library(downloader)  # download
library(tools)       # md5sum
```

## ggplot theme

[ggplot Themes Gallery](https://www.datanovia.com/en/blog/ggplot-themes-gallery/)

```{r ggplotTheme}
theme_set(theme_minimal() +

          theme(axis.text             = element_text(size = 10),
                axis.title            = element_text(size = 14),

                legend.position       = "bottom",

                plot.caption          = element_text(hjust = c(0.0,1.0),
                                                     size = 10),
                plot.caption.position = "plot",

                plot.title.position   = "plot",

                strip.background      = element_rect(fill = "aliceblue"),
                strip.text            = element_text(size = 14),

                title                 = element_text(size = 14)))

COLOR_BAR     <- "skyblue"
COLOR_OUTLINE <- "grey80"
```

## Helper functions

```{r Helpers}
Show <- function(data, caption="", bigMark="",
                 height = NULL, width = NULL, ...)
{
  data                                       |>
  kable("html", caption=caption,
        format.args=list(big.mark=bigMark))  |>
  kable_styling(bootstrap_options=c("striped", "bordered", "condensed"),
                position="left",
                full_width=FALSE, ...)       |>
  scroll_box(height = height, width = width)
}
```

# Read CVR

`Cecil.csv` was renamed from `cvr.csv` which was [downloaded from
Jeff O'Donnell's site](https://votedatabase.com/cvr/Maryland/2024/Cecil/)

Unclear where to find original information from [State of Maryland](https://elections.maryland.gov/press_room/index.html)

```{r}
cvr <- read_csv("Cecil.csv", 
                show_col_types = FALSE,
                guess_max = EXCEL_LIMIT)

glimpse(cvr)
```

# Summary of Key Variables

```{r}
cvr |>
  summarize(
            n          = n(),
            nUnique    = n_distinct(`Cast Vote Record`),
            nPrecinct  = n_distinct(Precinct),
            nStyle     = n_distinct(`Ballot Style`),
            nPresident = n_distinct(`President and Vice President of the United States (500001630)`) 
           )           |>
  Show()
```

In this case one `Ballot Style` per `Precinct`

# Gaps?

Do we care?

```{r}
table(diff(cvr$`Cast Vote Record`))
```

# Precinct runs

```{r}
precinctRun <- rle(cvr$Precinct)   # run-length encoding lengths
runs <- tibble(lengths = precinctRun$lengths, precinct = precinctRun$values)
glimpse(runs)
```

```{r}
table(runs$lengths)
```

Precincts with longest runs

```{r}
runs |> 
  filter(lengths> 740) |>
  arrange(lengths)     |>
  Show()
```


# Frequency Counts

## Presidential Candidates

```{r}
cvr |>
  group_by(`President and Vice President of the United States (500001630)`) |>
  count() |>
  Show()
```

## Precincts

### Presidential Vote

```{r}
presVote <- 
  cvr |>
  group_by(Precinct, 
           `President and Vice President of the United States (500001630)`)  |>
  count()                                                                    |>
  spread(`President and Vice President of the United States (500001630)`, n) |>
  ungroup()                                                                  |>
  select(Precinct, starts_with("Harris"), starts_with("Trump"))              |>
  rename(Harris = `Harris-Walz (10007007.0003640)`,
         Trump  = `Trump-Vance (10007007.0000670)`)

presVote |> head() |>  Show()

```

```{r}
presTotal <- 
  cvr |>
  group_by(Precinct) |>
  summarize(Voters = n()) |>
  inner_join(presVote, by = "Precinct")

presTotal |> Show()
```

### Ballot Style

```{r}
cvr |>
  group_by(Precinct, `Ballot Style`) |>
  count() |>
  Show()
```

# Compute CVR Vote Ratio

Use `coalesce` here to deal with NAs

Use largest `Cast Vote Record` even though there were gaps.

```{r}
cvrRecode <- 
  cvr |>
  mutate(
          Harris    = coalesce(as.integer(`President and Vice President of the United States (500001630)` == 
                                 "Harris-Walz (10007007.0003640)"),
                               0), 
          Trump     = coalesce(as.integer(`President and Vice President of the United States (500001630)` == 
                                 "Trump-Vance (10007007.0000670)"),
                               0),

          cumHarris = cumsum(Harris),
          cumTrump  = cumsum(Trump),
          VoteRatio = cumTrump / cumHarris,
          
          Percent   =  100 * `Cast Vote Record` / cvr$`Cast Vote Record`[nrow(cvr)]
        )         
```

# Plot CVR Vote Ratio

```{r CVR-Vote-Ratio, fig.width =10, fig.height = 8}
Stamp          <- paste("efg", format(time.1, "%Y-%m-%d"))      

ggplot(cvrRecode |> slice(-1),    # drop first row
       aes(x = Percent,y = VoteRatio)) + 
  geom_line(size = 2, color = "grey")       +
  scale_x_continuous(breaks = 0:10 * 10)    +
  scale_y_continuous(limits = c(0, 2.75),
                     breaks = 0:11 * 0.25)   +
  labs(
        title    = "Trump/Harris Vote Ratio - Cecil County, MD",
        subtitle = "Nov 2024 General Election",
        x = "Cast Vote Record Percent Processed",
        y = "Vote Ratio",    
        panel.grid.minor.y = element_blank(),
        caption = c("", Stamp)   
  )
```

# Explore problem 

Let's look at first 80% v last 20%

```{r}
problem <-
  cvrRecode |>
  filter(Percent >= 80,
         Percent <= 97.5)

dim(problem)
```

## First 80%

```{r}
firstVote <- 
  cvrRecode |>
   filter(Percent < 80)  |>
  
  group_by(Precinct, 
           `President and Vice President of the United States (500001630)`)  |>
  count()                                                                    |>
  spread(`President and Vice President of the United States (500001630)`, n) |>
  ungroup()                                                                  |>
  
  select(Precinct, starts_with("Harris"), starts_with("Trump"))              |>
  rename(First80Harris = `Harris-Walz (10007007.0003640)`,
         First80Trump  = `Trump-Vance (10007007.0000670)`)
```

## Last 20%

```{r}
lastVote <- 
  cvrRecode |>
   filter(Percent >= 80)  |>
  
  group_by(Precinct, 
           `President and Vice President of the United States (500001630)`)  |>
  count()                                                                    |>
  spread(`President and Vice President of the United States (500001630)`, n) |>
  ungroup()                                                                  |>
  
  select(Precinct, starts_with("Harris"), starts_with("Trump"))              |>
  rename(Last20Harris = `Harris-Walz (10007007.0003640)`,
         Last20Trump  = `Trump-Vance (10007007.0000670)`)
```

```{r}
presTotal <- 
  presTotal |>
  left_join(firstVote, by = "Precinct") |>
  left_join(lastVote, by = "Precinct")
```

```{r}
presTotal |> Show()
```

```{r}
write_xlsx(presTotal, "Presidential-Vote-by-Precinct.xlsx")
```

# Epilog {.tabset .tabset-fade .tabset-pills}

## {.active}

## Session Info

```{r devtoolsSessionInfo}
devtools::session_info()
```

</div>

```{r epilogDeltaTime, echo=FALSE}
time.2 <- Sys.time()
processingTime <- paste("Processing time:", sprintf("%.1f",
                        as.numeric(difftime(time.2,
                                            time.1, units="secs"))), "secs\n")
```

`r processingTime`
`r format(time.2, "%Y-%m-%d %H%M")`

